False Discovery Rates, A New Deal

### 2014-12-17 talks/ASC2014.newtext.Rmd
added some talks

@@

```{r setup, include=FALSE}
  #set global chunk options
  opts_chunk$set(cache=TRUE,autodep=TRUE,warning=FALSE)
  dep_auto()
```

@@

>>>>

---
title: "Test-Presentation"
author: "Matthew Stephens"
date: "September 23, 2014"
output: ioslides_presentation
---

@@

```{r setup, include=FALSE}
  #set global chunk options
  #opts_chunk$set(cache=TRUE,autodep=TRUE,warning=FALSE)
  #dep_auto()
```

@@

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with Plot

```{r, echo=FALSE}
plot(cars)
```

### 2014-12-17 talks/UNC2014.dev.Rmd
added some talks

---
title: "False Discovery Rates, A New Deal"
author: Matthew Stephens
date: November 21, 2014
output: ioslides_presentation
---

@@

@@

@@

- to the developers of **R**, **knitr**, **Rstudio** and **Pandoc**.

@@

- `ashr` package: `http://www.github.com/stephens999/ash`

## Pandoc Command used

`pandoc -s -S -i --template=my.beamer -t beamer -V theme:CambridgeUS -V colortheme:beaver  slides.md -o slides.pdf`

(alternative to produce html slides; but figures would need reworking)
`pandoc -s -S -i -t dzslides --mathjax slides.md -o slides.html`

Here is my session info:

>>>>

---
title: "False Discovery Rates, A New Deal"
author: Matthew Stephens
date: October 30, 2014
runtime: shiny
output: ioslides_presentation
---

@@

```{r, include=FALSE}
 #plot a histogram of z scores, highlighting the alternative distribution
#of z scores that is implied by pi0
nullalthist_z = function(z,pi0,nullcol="blue",altcol="cyan",...){
  h=hist(z, freq=FALSE,col=nullcol,nclass=40,ylim=c(-0.5,0.5),...)
  nb = length(h$breaks)
  nullexp = pi0 * (pnorm(h$breaks)[-1] - pnorm(h$breaks[-nb]))/(h$breaks[-1]-h$breaks[-nb])
  h$density = h$density - nullexp
  plot(h,add=TRUE,col=altcol,freq=FALSE)
}

 set.seed(100)
z = c(rnorm(5000,0,2),rnorm(5000,0,1))
```

## Interactive Illustration (shiny package)
```{r, echo=FALSE, cache=FALSE}
 library("shiny")
inputPanel(  
  sliderInput("pi0", label = "Proportion of nulls:",
              min = 0, max = 1, value = 0.5, step = 0.1)
)

renderPlot({
  nullalthist_z(z, input$pi0)
})
```

@@

## Coda: Modelling effects vs Standardized effects

- Recall we assumed $\beta_j | s_j \sim g(\cdot)$ independent of $s_j$.

- But it is plausible that distribution of $\beta_j$ depends on $s_j$.

- Might motivate assuming $z_j = \beta_j/s_j | s_j \sim g_1(\cdot)$.

- Indeed, this is the implicit assumption underlying standard procedures that rank by the $z$ scores.

- This model is equally easy to fit with same software! (And can compute log-likelihood of the different models.)

## Illustration: Modelling effects vs Standardized effects

```{r, echo=FALSE,fig.height=4,fig.cap="Model Effects",warning=FALSE,message=FALSE}
  plot(zdat[3,],abs(zdat[3,]/zdat[4,]),col=1+(zdat.ash$lfsr<0.05),xlim=c(-2,2),xlab="beta-hat",ylab="zscore")
```

## Illustration: Modelling effects vs Standardized effects

```{r, echo=FALSE,fig.height=4,fig.cap="Model Standardized Effects",warning=FALSE,message=FALSE}
  plot(zdat[3,],abs(zdat[3,]/zdat[4,]),col=1+(zdat.ash.ES$lfsr<0.05),xlim=c(-2,2),xlab="beta-hat",ylab="zscore")
```

Loglikelihood difference = `r zdat.ash$loglik-zdat.ash.ES$loglik` in favor of modelling effects.

@@

- to the developers of **R**, **knitr**, **shiny**, **Rstudio** and **Pandoc**.

@@

- `ashr` R package: `http://www.github.com/stephens999/ash`

## Session Info

### 2014-12-17 talks/gtex.Rmd
added some talks

---
title: "False Discovery Rates, A New Deal"
author: Matthew Stephens
date: October 30, 2014
runtime: shiny
output: ioslides_presentation
---

@@

## Before we get started: Let's get Organized!

- Over ~10 years of working with graduate students + postdocs,
I've noticed something.
- Organized researchers get more done (and better!).
- Many of them are more organized than I am!
- Thought: I should get organized; I should help others get organized.

## So what can you do?

- Buy a notebook; bring it to meetings; make notes! 
- Come to meetings with a written agenda. 
- While doing research, record what you did and what the outcome was.
- Use version control ([git](http://git-scm.com)) and internet repositories ([bitbucket](http://www.bitbucket.org), [github](http://www.github.com)) to organize notes, code, etc.
- Use *knitr* to help make your research reproducible.
- Talk about the tools you find useful!

## What are these repository things?

- A repository: a central place in which an aggregation of data is kept and maintained in an organized way (searcharticle.com)
- Great for sharing material across multiple people (eg student and advisor!)
- An amateur example: [http://github.com/stephens999/ash](http://github.com/stephens999/ash)

## What is knitr?

- An R package
- A tool for literate programming
- Text, and R code are interleaved
- When you compile the document, the code is run, and output inserted into the text.
- Great for writing reports, and keeping a track of what you did and what the result was!
- This talk was written with knitr (with RStudio)!

## What is Reproducible Research?

- Principle: when publishing results of computational procedures, we should
publish the code that produced the results.
- "publishing figures or results without the complete software environment could
be compared to a mathematician publishing an announcement of a mathematical theorem without giving the proof" (Buckheit and Donohoe)
- “an article about a computational result is advertising, not scholarship. The actual scholarship is the full software environment, code and data, that produced the result.” [Claerbout]

## Why is reproducibility important?

- Not only because people are forgetful, error-prone, or dishonest!
- Reproducing work is also the first step to extending it.
- Helps communications among researchers (eg student + advisor).
- If you do not publish code implementing your methods, your methods will likely go unused.

## More on git, github, knitr, reproducibility

- Google "The git book", to get started on git.
- Google "Karl Broman github tutorial" for statistics-oriented intro to github.
- Google "donohoe buckheit" for "Wavelab and reproducible research"

@@

## FDR, local fdr, and q values

Although precise definitions vary depending on whether one
takes a Bayesian or Frequentist approach to the problem, roughly

- The FDR at a threshold $P$ is 
$$\text{FDR}(P)=\Pr(\beta_j = 0 |  p_j<P).$$

- The q value for observation $j$ is $q_j=\text{FDR}(p_j)$.

- The local false discovery rate, fdr, at threshold $P$ is 
$$\text{fdr}(P) = \Pr(\beta_j =0 | p_j=P ).$$

- The fdr is more relevant, but slightly harder to estimate than 
FDR because it involves density estimation rather than tail-area estimation.

@@

## Is this an important problem?

- The original paper introducing FDR (Benjamini and Hochberg, 1995) has
been cited 21,787 times (May 2014) according to Google Scholar.

- That is three times a day for the last 19 years!

@@

## Mouse Data: Counts vary considerably across genes

```{r, echo=FALSE,fig.height=4,fig.cap=""}
hist(log10(xx),main="Distribution of total counts", xlab="log10(counts)")
```

@@

## Interactive Illustration (shiny package)
```{r, echo=FALSE, cache=FALSE}
 library("shiny")
inputPanel(  
  sliderInput("pi0", label = "Proportion of nulls:",
              min = 0, max = 1, value = 0.5, step = 0.1)
)

renderPlot({
  nullalthist_z(z, input$pi0)
})
```

@@

- In the illustrative example, the maximum $q$ value is `r round(max(hh.ash$qvalue),2)`

@@

## Advantages of False Sign Rate

- The FDR is typically not actually identifiable from data.

- This is because data cannot distinguish between $\beta_j = 0$ and $\beta_j$ "very
small". So $\pi_0$ is not identifiable, and FDR is very sensitive to $\pi_0$.

- So methods for estimating $\pi_0$ and FDR, including those presented here, are designed to be ``conservative" (i.e. overestimate the FDR).

- The False Sign Rate is much less senstive to $\pi_0$, and hence more identifiable from data!

## Simulated example: $\pi_0$ not identifiable.

```{r,echo=FALSE,include=FALSE}
load("../paper/Rcode/sim1.RData") #load simulation results
source("../paper/Rcode/plot_pi0.R")
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
plot_pi0(list(simres1a))
```

## Simulated Example: so fdr not identifiable

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
source("../paper/Rcode/plot_lfsr.R")
  plot_lfsr(list(simres1a),0.1,ptype="lfdr")
```

## Simulated Example: fsr much more identifiable

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  plot_lfsr(list(simres1a),0.1,ptype="lfsr")
```

@@

## Coda: Modelling effects vs Standardized effects

- Recall we assumed $\beta_j | s_j \sim g(\cdot)$ independent of $s_j$.

- But it is plausible that distribution of $\beta_j$ depends on $s_j$.

- Might motivate assuming $z_j = \beta_j/s_j | s_j \sim g_1(\cdot)$.

- Indeed, this is the implicit assumption underlying standard procedures that rank by the $z$ scores.

- This model is equally easy to fit with same software! (And can compute log-likelihood of the different models.)

## Illustration: Modelling effects vs Standardized effects

```{r, echo=FALSE,fig.height=4,fig.cap="Model Effects",warning=FALSE,message=FALSE}
  plot(zdat[3,],abs(zdat[3,]/zdat[4,]),col=1+(zdat.ash$lfsr<0.05),xlim=c(-2,2),xlab="beta-hat",ylab="zscore")
```

## Illustration: Modelling effects vs Standardized effects

```{r, echo=FALSE,fig.height=4,fig.cap="Model Standardized Effects",warning=FALSE,message=FALSE}
  plot(zdat[3,],abs(zdat[3,]/zdat[4,]),col=1+(zdat.ash.ES$lfsr<0.05),xlim=c(-2,2),xlab="beta-hat",ylab="zscore")
```

Loglikelihood difference = `r zdat.ash$loglik-zdat.ash.ES$loglik` in favor of modelling effects.

## Other Applications

- Widely applicable: requiring only an estimated
effect size and standard error for each object.

- E.g. Currently applying it to wavelet shrinkage applications.

@@

- NHGRI for funding.

>>>>

---
title: "False Discovery Rates, A New Deal"
author: "Matthew Stephens"
date: "November 24, 2014"
output: ioslides_presentation
---

@@

@@

@@

@@

@@

@@

- Actually this can be true for existing `qvalue` approach too!

@@

@@

@@

- NHGRI and GTEX for funding.
